Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2017-12-28 paper-ieee-trans.tex
Update paper-ieee-trans.tex

\subsection{INtegrating Distributed data Infrastructures for Global
ExplOitation}

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
ExplOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
the project is succeeding in providing solutions to address the existing gaps in
cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

The INDIGO--Datacloud project reflects the progress made in software
quality and reliability aspects throughout the past scientific European
experiences. Nevertheless, the project’s Quality Assurance procedures are also
highly influenced by the insights of current big worldwide collaborations of
software development. These collaborations have continuously inspired the use new
technologies and procedures to increase the robustness and quality of the
software being delivered. In this scenario, the implementation of the DevOps culture 
is one of the major outcomes, as this paradigm has been progressively adopted by
project.

@@

The DevOps methods emphasize on exercising Quality Assurance techniques to
avoid infrastructure disruption whenever new developments are deployed into
production. The INDIGO--DataCloud project promoted the application of a Continuous Integration (CI)
scenario to enforce the Quality Assurance requirements for any piece of
software being produced within the project. Such environment requires an
automated ready--to--go infrastructure where the different services involved
(source code management platform, automation server) interact with each other
to trigger the quality check execution and return back the exit codes, together
with their outputs. To accomplish this, the project leveraged on tightly
integrated open source tools such as GitHub and Jenkins, as well as relying on
Docker container provisioning to speed up the source code validation. This CI
approach is guiding the project’s software development phase throughout the
first and second major releases.

As DevOps suggests, frequent releases positively affects the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the execution of
the quality checks (as part of the CI). 
The steps defined within the CD
pipeline differ whether the software is to be distributed via Docker
images (see Figure \ref{fig:fig_CD}) or via operating system’s packages. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
built in the previous step and uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}.

@@

The software quality procedures \cite{indigo-d31} cover: 1) the identification
and description of the \emph{quality requirements} that the software need to
comply with, and 2) the \emph{quality metrics}.

The requirements are applied at early stages in the software development
process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
bug or design issue is likely to be detected and corrected promptly. The source
code is publicly available in GitHub repositories under an organization called
{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the product
catalogue, promoting the external contributions and software adoption. Furthermore,
the source code is compliant with community de--facto style standards,
selected from the variety of programming languages being used as seen in
Figure \ref{fig:fig_codestyle}.

@@

Unit testing coverage has increased between the first ({\sl INDIGO-1}) and the
second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
66.69\% on the {\sl INDIGO-2} release, very close to the recommended 70\% threshold
defined by the project’s SQA base criteria \cite{indigo-d31}. A comparison graph
between the code coverage for the products involved both in {\sl INDIGO-1} and
{\sl INDIGO-2} releases is shown in Figure \ref{fig:fig_unittest}. As it can be seen,
a very small set of products with low coverage values highly impacted the overall
result, although a significant part of these products are upstream contributions
or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them have values above
the 50\% coverage.

The documentation of the products is treated as source code, using a markup
language, automatically rendered and uploaded to online repositories
\cite{indigo-gitbook}. Changes in both documentation and source code are
human--reviewed as the last step before merging them into the production
repository.

Last but not least, as described in section \ref{sec:devops}, in order
to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
automatically using either Ansible \cite{indigo-ansible} or Puppet
\cite{indigo-puppet} open--source tools. The adoption of such Configuration Management
tools not only eases the consumption of the products developed by external users or
infrastructures, but also tackles the correct installation and configuration of the
services and applications. A representative example is the number of roles currently
hosted in the Ansible Galaxy portal for the project.

@@

Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
for deployment. It shows an increase after the {\sl INDIGO-1} release with specially
in the previous weeks of each of the two major releases.

@@

Two applications from WP2, DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is deployed as a Docker container, and subsequently uploaded
to the Dockerhub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
Thus this pipeline implementation goes a step forward by testing the application
execution for the last available Docker image in the catalogue.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the INDIGO--DataCloud
solution and use cases validation, by deploying and executing the applications
with the last stable version of the software. A map of the Pilot Preview
infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.

The released software is also tested in production environments through the
stage--rollout process. Selected resource providers are requested to install
the most updated stable versions of the software, accessed by their users. The
stage--rollout process is key to detect and mitigate issues that could only
appear in production.

@@

It is a fact that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. Recent software engineering insights, such as
DevOps, are gradually being applied as the open--source collaborative tools
evolve, allowing tighter integrations. The focus of the Software Quality
Assurance procedures has been integrated in the development stage, trying to
detect and correct issues early in the software lifecycle. Automation and
metric analysis are at the base of prompt issue solving. However, post--release
validation, such as integration testbeds and the stage--rollout process, are
also needed to strengthen the software reliability for scientific usage.

>>>>

\subsection{INtegrating Distributed data Infrastructures for Global
explOitation}

The INDIGO-DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
the project is succeeding in providing solutions to address the existing gaps in
cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

The INDIGO--Datacloud project reflects the progress made in software
quality and reliability aspects throughout the past scientific European
experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
highly influenced by the insights of current big worldwide collaborations of
software development. These collaborations have continuously inspired the use of new
technologies and procedures to increase the robustness and quality of the
software being delivered. In this scenario, the implementation of the DevOps culture 
is one of the major outcomes, as this paradigm has been progressively adopted by the
project.

@@

The DevOps methods emphasize on exercising Quality Assurance techniques to
avoid infrastructure disruption whenever new developments are deployed into
production. The INDIGO--DataCloud project promoted the application of a 
Continuous Integration (CI) scenario to enforce the Quality Assurance requirements 
for any piece of software being produced within the project. Such environment 
requires an automation ready--to--go infrastructure where the different services involved
(source code management platform, automation server) interact with each other
to trigger the quality check execution and return back the exit codes, together
with their outputs. To accomplish this, the project leveraged on tightly
integrated open source tools such as GitHub [ref], as source code repository
and Jenkins [ref], as continuous integration system, as well as relying on
Docker container provisioning to speed up the source code validation. This CI
approach is guiding the project’s software development phase throughout the
first and second major releases.

As DevOps suggests, frequent releases positively affect the reliability of the
software. The software updates of INDIGO--DataCloud products taking place since
the second major release are passing through a Continuous Delivery (CD)
pipeline that adds the packaging of the software right after the successful execution of
the quality checks (as part of the CI). 
The steps defined within the CD pipeline differ whether the software is to be distributed via Docker
images (see Figure \ref{fig:fig_CD}) or via operating system’s packages, rpms or debs. In the latter case an extra
validation step is added, consisting in the product’s deployment using a
Configuration Management (CM) solution. The installation uses the packages
built in the previous step and uploaded to a testing
repository. In the case of Docker images, the CM is used to build the image
itself, installing and configuring the product before uploading it to the
DockerHub repository \cite{indigo-dockerhub}.

@@

The software quality policies \cite{indigo-d31} cover: 1) the identification
and description of the \emph{quality requirements} that the software need to
comply with, and 2) the \emph{quality metrics}.

The requirements are checked at early stages in the software development
process, integrated in the Jenkins CI service \cite{indigo-jenkins}, so that any
bug or design issue is likely to be detected and corrected promptly. The source
code is publicly available in GitHub repositories under the project organization called
{\sl indigo-dc} \cite{indigo-github} to increase the visibility of the products
catalogue, promoting the external contributions and software adoption. Furthermore,
the source code is checked for compliance with community de--facto style standards,
selected for each one of the variety of programming languages being used as seen in
Figure \ref{fig:fig_codestyle}.

@@

Unit testing coverage, another project quality metric, has increased between the first ({\sl INDIGO-1}) and the
second ({\sl INDIGO-2}) INDIGO--DataCloud releases, reaching an average value of
66.69\% on the {\sl INDIGO-2} release, very close to the recommended 70\% threshold
defined by the project’s SQA base criteria \cite{indigo-d31}. A comparison graph
between the code coverage for the products involved both in {\sl INDIGO-1} and
{\sl INDIGO-2} releases is shown in Figure \ref{fig:fig_unittest}. As it can be seen,
a very small set of products with low coverage values highly impacted the overall
result, although a significant part of these products are upstream contributions
or products with codebases not fully under the control of the project. Nevertheless,
60\% of the products were over this threshold while 80\% of them had values above
the 50\% coverage.

The documentation of the products is treated as source code, using a markup
language, automatically rendered and uploaded to online repositories
\cite{indigo-gitbook}. Changes in both documentation and source code are
human--reviewed as the last step before merging them into the production
repository or official documentation.

Last but not least, as described in section \ref{sec:devops}, in order
to facilitate the usage of the INDIGO--DataCloud services, they can be deployed
automatically using either Ansible \cite{indigo-ansible} or Puppet
\cite{indigo-puppet} open--source configuration tools. The adoption of such Configuration Management
tools not only eases the consumption of the products developed by external users or
infrastructures, but also tackles the correct installation and configuration of the
services and applications. A representative example is the number of roles currently
hosted in the Ansible Galaxy portal for the project [ref].

@@

Figure \ref{fig:fig_confman} shows the number of products that offer an automated means
for deployment. It shows an increase after the {\sl INDIGO-1} release especially
in the previous weeks of each of the two major releases.

@@

Two applications from "Definition of Support to Research Communities" workpackage, 
DisVis \cite{disvis} and PowerFit \cite{powerfit}, were
integrated into a similar CI/CD pipeline described in section \ref{sec:devops}.
Figure \ref{fig:fig_disvis} shows the pipeline for the Disvis application.

@@

The novelty introduced in the pipeline above is the validation of the application.
Once the application is packaged as a Docker image, and subsequently uploaded
to the Dockerhub repository, it is instantiated in a new container to be validated.
The application is then executed and the results compared with a set of reference outputs.
Thus this pipeline implementation goes a step forward by testing the application
execution for the latest available Docker image in the catalogue.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the INDIGO--DataCloud
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as 
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.

The released software is tested in production environments through the
Staged--Rollout process, before inclusion in the EGI's CloudManagement Distribution (CMD). 
Selected resource providers are requested to install
the most updated stable versions of the software, and to give access to them to their users. The
Staged--Rollout process is key to detect and mitigate issues that could only
appear in production environments.

@@

It is a fact that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. Recent software engineering insights, such as
DevOps approaches, are gradually being applied as the open--source collaborative tools
evolve, allowing tighter integrations. The Software Quality
Assurance procedures have been integrated in the development stage, in order to
detect and correct issues early in the software lifecycle. Automation and
metrics analysis are at the base of prompt issue solving. However, post--release
validation, through the preview testbeds or the Staged--rollout process, are
also needed to strengthen the software reliability for scientific usage.
